Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-29 pnas/Towards.tex
Two Gaussians proof

In general, for a signal $z$ of length $m$, the autocorrelation of order $q \geq 1$ is given for any integer shifts $\ell_1, \ldots, \ell_{q-1}$ by
\begin{align}
a_z^q[\ell_1,\ldots,\ell_{k-1}]  & = \frac{1}{m} \sum_{i=-\infty}^{+\infty} z[i]z[i+\ell_1]\cdots z[i+\ell_{q-1}],
\label{eq:ac_general}
\end{align}
where indexing of $z$ out of the range $0, \ldots, m-1$ is zero-padded.
For our purposes, this will be applied both to $x$ (of length $L$) and to $y$ (of length $N$).
Explicitly, the first-, second- and third-order autocorrelations are given by \TODO{we may want to remove these explicit formulas}
\begin{align} 
a_z^1 & = \frac{1}{m} \sum_{i=0}^{m-1} z[i], \nonumber\\
a_z^2[\ell] & = \frac{1}{m} \sum_{i = \max\{0, -\ell\}}^{m-1 + \min\{0, -\ell\}} z[i]z[i+\ell], \nonumber\\
a_z^3[\ell_1,\ell_2] & = \frac{1}{m} \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{m-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. \label{eq:ac_special}
\end{align}
The autocorrelation functions have symmetries. Specifically, $a_z^2[\ell] = a_z^2[-\ell]$, and
$a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].
$
%\begin{align*}
%a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].
%\end{align*}
%Taking these symmetries into consideration, one can show that $a_z^1, a_z^2$ and $a_z^3$ contain \TODO{TKTK--I think it was commented in the latex} (respectively) non-trivial distinct real numbers, generically.
%
%For the special case $K = 1$ where a single signal $x = x_1$ must be recovered, the relation between autocorrelations of the micrograph and those of $x$ is particularly simple, so that we treat it first. It is useful to introduce some notation: let $M$ denote the number of occurrences of $x$ in $y$, and let
%\begin{align}
%\gamma & = \frac{M L}{N}
%\end{align}
%denote the density of $x$ in $y$ (that is, the fraction of entries of $y$ occupied by occurrences of $x$.) The spacing constraint~\eqref{eq:spacing} imposes $\gamma\leq\frac{L}{2L-1}\approx 1/2$.

@@

If $x$ is known, then the support signal $s$ can be estimated via linear programming  in the high SNR regime~\cite{azais2015spike,denoyelle2017support,bendory2016robust,bendory2017robust}. However, in the low SNR regime---even if $x$ is known---estimating the binary sparse signal $s$ is impossible. 
\TODO{Here we need to provide Will's argument as a proposition}

%We have few evidences for that\cite{aguerrebere2016fundamental}, but we will provide Will's argument here.

Accordingly, we must consider methods that aim to estimate the signal directly, without estimating the nuisance variables as an intermediate step. 
 %we need to consider a different approach. In this paper, we consider autocorrelation analysis.
  In what follows, we present several results on properties of autocorrelations. For simplicity, we consider one-dimensional signals, but the results carry through for higher dimensions.

@@

Some remarks are in order. First, \eqref{eq-uniqueness} is not numerically stable if $z[0]$ or $z[L-1]$ are close to 0. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares (LS) procedure, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.
Second, if the separation condition holds, then the length of the signal can be determined from the autocorrelations.
%%%and therefore the assumption that the first and last entries are nonzero is met.
In particular, if the separation condition holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Third, note that the second-order autocorrelation is not by itself sufficient to determine the signal uniquely. %~\cite{beinert2015ambiguities,bendory2017fourier}.
However, for dimensions greater than one, almost all signals are determined uniquely up to sign (phase for the complex signals) and reflection through the origin (with conjugation in the complex case). %~\cite{hayes1982reconstruction,hayes1982reducible}. 
The sign ambiguity can be resolved by the mean of the signal if it is not zero. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.

@@

Using third-order autocorrelation information of $y$, both the ratio $\gamma$ and the noise $\sigma$ are identifiable. For the following results, when we say that a result holds for a ``generic'' signal $x$, we mean that the set of signals which cannot be determined by these measurements
lies in a set with Lebesgue measure zero. 
%in the vanishing locus of a nonzero polynomial\footnote{Given a  nonzero polynomial $f$ in $T$ variables (real or complex), the vanishing locus of $f$ is the set 
%	$V(f) = \{(x_1, \ldots, x_T) \in \mathbb{R}^T (\text{resp. } \mathbb{C}^T) \vert f(x_1,\ldots x_T)=0\}$. By a general result in algebraic geometry $\dim V(f) < T$. As a consequence the complement of $V(f)$ in $\mathbb{R}^T$ (resp. $\mathbb{C}^T$) is dense.}. 
In particular, this means that we can recover
almost all signals with the given measurements.
%%
%we mean that it holds for all $x$ inside a set $\Omega \subset \RL$, whose complement $\RL \setminus \Omega$ has Lebesgue measure zero.
%
\begin{proposition} \label{prop:gamma_sigma}
	%
	%Let $\sigma > 0$ be fixed. 
	The observed autocorrelations $a_y^1,a_y^2$ and  $a_y^3$ determine the ratio $\gamma$ and noise level $\sigma$ uniquely for a generic signal $x$. If $\gamma\geq\frac{L}{4(L-1)}\approx \frac{1}{4}$, then this holds for any signal $x$ with nonzero mean. 
	\begin{proof}
		See Appendix~\ref{sec:proof_prop_gamma_sigma}.
	\end{proof}
\end{proposition}

@@

@@

\section{Connection to volume}

>>>>

In general, for a signal $z$ of length $m$, the autocorrelation of order $q \geq 1$ is given for any integer shifts $\ell_1, \ldots, \ell_{q-1}$ by
\begin{align}
a_z^q[\ell_1,\ldots,\ell_{k-1}]  & = \frac{1}{m} \sum_{i=-\infty}^{+\infty} z[i]z[i+\ell_1]\cdots z[i+\ell_{q-1}],
\label{eq:ac_general}
\end{align}
where indexing of $z$ out of the range $0, \ldots, m-1$ is zero-padded.
For our purposes, this will be applied both to $x$ (of length $L$) and to $y$ (of length $N$).
Explicitly, the first-, second- and third-order autocorrelations are given by \TODO{we may want to remove these explicit formulas}
\begin{align} 
a_z^1 & = \frac{1}{m} \sum_{i=0}^{m-1} z[i], \nonumber\\
a_z^2[\ell] & = \frac{1}{m} \sum_{i = \max\{0, -\ell\}}^{m-1 + \min\{0, -\ell\}} z[i]z[i+\ell], \nonumber\\
a_z^3[\ell_1,\ell_2] & = \frac{1}{m} \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{m-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. \label{eq:ac_special}
\end{align}
The autocorrelation functions have symmetries. Specifically, $a_z^2[\ell] = a_z^2[-\ell]$, and
$a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].
$

@@

If $x$ is known, then the support signal $s$ can be estimated via linear programming  in the high SNR regime~\cite{azais2015spike,denoyelle2017support,bendory2016robust,bendory2017robust}. 
However, in the low SNR regime---even if $x$ is known---estimating the binary sparse signal $s$ is impossible. 
To prove this argument, we consider an even easier problem, in which someone hands us intervals of length $L$ that either contain the full signal plus noise, or contain just noise, and we are asked to determine which intervals contain signal. 
Let us assume too that we know the signal $x$, that we know the probability $q$ that one of the intervals contains a signal, and that we know the noise variance $\sigma^2$. Since we have the full generative model, we can always generate more data as needed; therefore it is enough to consider the hardness of deciding whether a single interval contains signal or not.

Let us abstract this problem. We have two known vectors $\theta_0$ and $\theta_1$ in $\mathbb{R}^L$. There is a random variable $\eta$ taking values 0 or 1 with probabilities $q$ and $1-q$, respectively. We observe a random vector $X \in \mathbb{R}^L$ with the following distribution: $X | (\eta  = i) \sim N(\theta_i,\sigma^2)$; in other words, if $\eta = 0$, then $X \sim N(\theta_0,\sigma^2)$, and if $\varepsilon  = 1$, then $X \sim N(\theta_1,\sigma^2)$. If $\theta_1=0$, then the latter corresponds to pure noise interval.

We observe $X$, and our task is to predict $\eta$. How well can we do it? First, let us assume without loss of generality that $q \ge 1/2$. Then, if we make the constant prediction $\hat{\eta } \equiv 0$, we will be correct with probability $q$. So the question is, can we do better than this? We will prove that as $\sigma \to \infty$, the answer is no. More precisely:
\begin{proposition} \label{prop:two_gauss}
	Let $\hat{\eta }$ be any predictor of $\eta$.
	%
	\begin{align*}
	%
	\lim_{\sigma \to \infty} \P[ \hat{\eta} = \eta  ] \le q.
	%
	\end{align*}
\end{proposition}
\noindent The result is proved in Appendix~\ref{sec:proof_two_gauss}.

Proposition~\ref{prop:two_gauss} implies that in order to estimate the signal in low SNR  we must consider methods that aim to estimate the signal directly, without estimating the nuisance variables as an intermediate step. 
 %we need to consider a different approach. In this paper, we consider autocorrelation analysis.
  In what follows, we present several results on properties of autocorrelations. For simplicity, we consider one-dimensional signals, but the results carry through for higher dimensions.

@@

Some remarks are in order. First, \eqref{eq-uniqueness} is not numerically stable if $z[0]$ or $z[L-1]$ are close to 0. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares (LS) procedure, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.
Second, if the separation condition holds, then the length of the signal can be determined from the autocorrelations.
In particular, if the separation condition holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Third, note that the second-order autocorrelation is not by itself sufficient to determine the signal uniquely. %~\cite{beinert2015ambiguities,bendory2017fourier}.
However, for dimensions greater than one, almost all signals are determined uniquely up to sign (phase for the complex signals) and reflection through the origin (with conjugation in the complex case). %~\cite{hayes1982reconstruction,hayes1982reducible}. 
The sign ambiguity can be resolved by the mean of the signal if it is not zero. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.

@@

Using third-order autocorrelation information of $y$, both the ratio $\gamma$ and the noise $\sigma$ are identifiable. For the following results, when we say that a result holds for a ``generic'' signal $x$, we mean that the set of signals which cannot be determined by these measurements
lies in a set with Lebesgue measure zero. 
In particular, this means that we can recover
almost all signals with the given measurements.

\begin{proposition} \label{prop:gamma_sigma}
	%
	%Let $\sigma > 0$ be fixed. 
	The observed autocorrelations $a_y^1,a_y^2$ and  $a_y^3$ determine the ratio $\gamma$ and noise level $\sigma$ uniquely for a generic signal $x$. If $\gamma\geq\frac{L}{4(L-1)}\approx \frac{1}{4}$, then this holds for any signal $x$ with nonzero mean. 
	\begin{proof}
		See Appendix~\ref{sec:proof_prop_gamma_sigma}.
	\end{proof}
\end{proposition}

@@

\section{Proof of Proposition~\ref{prop:two_gauss}} \label{sec:proof_two_gauss}

The proof is based on  a variant of the Neyman-Pearson Lemma to derive the best (deterministic) predictor $\hat{\eta}$. We take any predictor $\hat{\eta}$, and we let $S$ be the set of $X$'s where $\hat{\eta} = 1$. Then the probability that $\hat{\eta}$ fails is:
%
\begin{align}
\label{eq:neyman-pearson}
%
\P[\hat{\eta} \ne \eta] 
&= q \P_0[\hat{\eta} = 1 ] + (1-q) \P_1[\hat{\eta} = 0]
\nonumber \\
&= q \P_0[\hat{\eta} = 1] + (1-q) (1 - \P_1[\hat{\eta} = 1])
\nonumber \\
&= q \P_0[\hat{\eta} = 1] + (1-q) - (1-q) \P_1[\hat{\eta} = 1]
\nonumber \\
&= (1-q)  + \int_S (q f_0(x) - (1-q) f_1(x))dx,
%
\end{align}
%
where $f_i(x)$ is the normal density with mean $\theta_i$ and variance $\sigma^2$, and the notation $\P_i$ means probability conditional on the event $\eta = \theta_i$; that is, $\P_i[A] = \P[A | \eta=i]$. 
Let $\Lambda(x) = f_0(x) / f_1(x)$ and $b = (1-q) / q$,
The best predictor of $\eta$ based on $X$ will minimize the integral~\eqref{eq:neyman-pearson} and thus \TODO{Tamir: Are we sure this is the best?}
%
% is minimized by taking $S$ to be the set:
%%
%\begin{align}
%%
%S = \left\{x : \frac{f_0(x)}{f_1(x)} \le \frac{1-q}{q} \right\}.
%%
%\end{align}
%%
%In other words, if $\Lambda(x) = f_0(x) / f_1(x)$ and $b = (1-q) / q$, then the best predictor of $\eta$ based on $X$ is:
%
\begin{align}
%
\eta = 
\begin{cases}
1,  \text{ if } \Lambda(x) \le b \\
0,  \text{ if } \Lambda(x) \ge b
\end{cases}.
%
\end{align}

Taking logarithms, the set $S$ can be rewritten as the set of $x$ where:
\begin{align*}
%
-\| x - \theta_0\|^2 \le - \|x - \theta_1\|^2 + 2 \sigma^2 \log(b),
%
\end{align*}
%
or equivalently
%
\begin{align*}
%
\langle x , \theta_1 - \theta_0 \rangle 
\ge \frac{\|\theta_1\|^2 - \|\theta_0\|^2}{2} - \sigma^2 \log(b).
%
\end{align*}
%

Now let us compute the probability of failure conditional on the event $\eta = 0$. In this case, failure occurs when $X \in S$. Since $X|(\eta=0) \sim N(\theta_0,\sigma^2)$, we can write $X | (\eta=0) = \sigma Z + \theta_0$, where $Z \sim N(0,1)$. Consequently,
%
\begin{align*}
%
\langle X , \theta_1 - \theta_0 \rangle 
&= \sigma \langle Z , \theta_1 - \theta_0 \rangle 
+ \langle \theta_0,\theta_1 - \theta_0 \rangle
\nonumber \\
&= \sigma \langle Z , \theta_1 - \theta_0 \rangle 
+ \langle \theta_0,\theta_1 \rangle - \|\theta_0\|^2,
%
\end{align*}
%
and failure occurs when 
%
\begin{align}
%
\sigma\langle Z, \theta_1 - \theta_0 \rangle 
&\ge \frac{\|\theta_1\|^2 + \|\theta_0\|^2}{2} - \langle \theta_0,\theta_1 \rangle
- \sigma^2 \log(b)
\nonumber \\
&= \frac{1}{2} \|\theta_1 - \theta_0\|^2 - \sigma^2 \log(b)
%
\end{align}
%
or equivalently
%
\begin{align*}
%
Y \ge \frac{c}{\sigma} - \sigma^2 \log(b)
%
\end{align*}
%
where $c = \|\theta_1 - \theta_0\|^2 / 2$ and $Y = \langle Z , \theta_1 - \theta_0 \rangle \sim N(0,\|\theta_1 - \theta_0\|^2)$. 
Just for simplicity let us assume $\| \theta_1 - \theta_0\| = 1$, so that $Y \sim N(0,1)$. So:
%
\begin{align*}
%
\P_0[\hat{\eta} = 1] = \P\left[Y \ge \frac{c}{\sigma} - \sigma^2 \log(b) \right],
\quad Y \sim N(0,1).
%
\end{align*}
%
Similarly,
%
\begin{align*}
%
\P_1[\hat{\eta} = 0] = \P\left[Y \ge \frac{c}{\sigma} + \sigma^2 \log(b) \right],
\quad Y \sim N(0,1).
%
\end{align*}
%
So the overall probability of failure is:
%
\begin{align*}
%
\P[\hat{\eta} \ne \eta]
&= q \P\left[Y \ge \frac{c}{\sigma} - \sigma^2 \log(b) \right]
\\&+ (1-q) \P\left[Y \ge \frac{c}{\sigma} + \sigma^2 \log(b) \right].
%
\end{align*}
%
Now, if $q = 1/2$, then $\log(b) = 0$. Hence the probability of failure is simply:
%
\begin{align}
%
\P\left[Y \ge \frac{c}{\sigma} \right] \longrightarrow \frac{1}{2} = q
\quad \text{as} \quad \sigma \to \infty.
%
\end{align}

On the other hand, if $q > 1-q$, then $\log(b) < 0$. Then 
%
\begin{align}
%
\P\left[Y \ge \frac{c}{\sigma} - \sigma^2 \log(b) \right] \longrightarrow 0
%
\end{align}
%
while
%
\begin{align}
%
\P\left[Y \ge \frac{c}{\sigma} + \sigma^2 \log(b) \right] \longrightarrow 1
%
\end{align}
%
as $\sigma \to \infty$. Consequently,
%
\begin{align}
%
\P[\hat{\eta} \ne \eta]
&= q \P\left[Y \ge \frac{c}{\sigma} - \sigma^2 \log(b) \right]
+ (1-q) \P\left[Y \ge \frac{c}{\sigma} + \sigma^2 \log(b) \right]
\nonumber\\
& \longrightarrow 1-q  \quad \text{as} \quad  \sigma \to \infty.
%
\end{align}
%
That is, the probability of success converges to $q$, which completes the proof.

@@

\subsection{Connection to volume}
